---
title: "R Notebook for Climate Change Trends"
output: html_notebook
---

Loading the libraries requied to analyze the datasets

```{r}
#loading the required packages
library(tidyverse)
library(corrplot)
library(VIM)
library(Hmisc)
library(psych)
library(ggplot2)
library(dplyr)
library(plotly)
library(GGally)
library(mice)
library(readr)

```

Loading the datasets

```{r}
#loading datasets
df1 <- read_csv("global-temp-annual.csv")
df2 <- read_csv("owid-co2-data.csv")

```

Summary of dataset

```{r}
#summary of datasets
cat("Dataset 1 Dimensions:", dim(df1), "\n")
cat("Dataset 2 Dimensions:", dim(df2), "\n")
cat(strrep("-", 50), "\n")
cat("Summary of Dataset 1: \n")
summary(df1)
cat(strrep("-", 50), "\n")
cat("Summary of Dataset 2: \n")
summary(df2)
cat(strrep("-", 50), "\n")
```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
